Competing Bandits:\\ The Perils of Exploration under Competition \thanks{This is a merged and final version of two conference papers, \citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19}, with a unified and streamlined presentation.}

### 2020-07-08 journal paper/content/related_work.tex
tweaks to paper
updated related work + intro

The literature on MAB is vast and multi-threaded. The most related
thread concerns regret-minimizing MAB formulations with IID rewards
\citep{Lai-Robbins-85,bandits-ucb1}. This thread includes ``smart" MAB
algorithms that combine exploration and exploitation, such as UCB1
\citep{bandits-ucb1} and Successive Elimination
\citep{EvenDar-icml06}, and ``naive'' MAB algorithms that separate
exploration and exploitation, including explore-first and
$\eps$-Greedy \citep[\eg see][]{slivkins-MABbook}.

@@

Our setting is closely related to the ``dueling algorithms" framework \citep{DuelingAlgs-stoc11} which studies competition between two principals, each running an algorithm for the same problem. However, this work considers algorithms for offline / full input scenarios, whereas we focus on online machine learning and the explore-exploit-incentives tradeoff therein. Also, this work specifically assumes binary payoffs (\ie win or lose) for the principals.

@@

A line of work on \emph{platform competition}, starting with \cite{Rysman09}, concerns competition between firms (\emph{platforms}) that improve as they attract more users (\emph{network effect}); see \citet{Weyl-White-14} for a recent survey. This literature is not concerned with \innovation, and typically models network effects exogenously, whereas in our model network effects are endogenous: they are created by MAB algorithms, an essential part of the model.

Relaxed versions of rationality similar to ours are found in several notable lines of work. For example, ``random agents" (a.k.a. noise traders) can side-step the ``no-trade theorem'' \citep{Milgrom-Stokey-82}, a famous impossibility result in financial economics. The \SoftMaxRandom model is closely related to the literature on \emph{product differentiation}, starting from \cite{Hotelling-29}, see \cite{Perloff-Salop-85} for a notable later paper.

There is a large literature on non-existence of equilibria due to small deviations   (which is related to the corresponding result for \HardMaxRandom), starting with \cite{Rothschild-Stiglitz-76} in the context of health insurance markets. Notable recent papers \citep{Veiga-Weyl-16,Azevedo-Gottlieb-17} emphasize the distinction between \HardMax and versions of \SoftMaxRandom.

>>>>

The literature on MAB is vast and multi-threaded. The most related
thread concerns regret-minimizing MAB formulations with IID rewards
\citep{Lai-Robbins-85,bandits-ucb1}. This thread includes ``smart" MAB
algorithms that combine exploration and exploitation, such as UCB1
\citep{bandits-ucb1}, Thompson Sampling \citep{russo2018tutorial}, and Successive Elimination
\citep{EvenDar-icml06}, and ``naive'' MAB algorithms that separate
exploration and exploitation, including explore-first and
$\eps$-Greedy \citep[\eg see][]{slivkins-MABbook}.

@@

Our setting is closely related to the ``dueling algorithms" framework \citep{DuelingAlgs-stoc11} which studies competition between two principals, each running an algorithm for the same problem. However, this work considers algorithms for offline / full input scenarios, whereas we focus on online machine learning and the explore-exploit-incentives tradeoff therein. Also, this work specifically assumes binary payoffs (\ie win or lose) for the principals. \gaedit{\cite{ben2017best, ben2019regression} similarly study competing learning algorithms, though they consider competition between offline learning algorithms, whereas our focus here is on online learning algorithms.}

@@

\gaedit{There is a literature that studies the interaction between learning-by-doing and competition \citep{fudenberg1983learning, dasgupta1988learning, cabral1994learning}. In these models, firms learn while competing against each other where a firm attracting more consumers reduces its per-unit production costs. However, our model focuses on firms learning product quality as opposed to reducing production costs and the impact that this learning has on attracting consumers.}

\gaedit{A line of work on \emph{platform competition}, starting with \cite{Rysman09}, concerns competition between firms (\emph{platforms}) that improve as they attract more users (\emph{network effect}); see \citet{Weyl-White-14} for a recent survey. This literature is not concerned with \innovation, and typically models network effects exogenously, whereas in our model network effects are endogenous: they are created by MAB algorithms, an essential part of the model. A nascent literature studies whether and when network effects manifest themselves in data-intensive markets \citep{prufer2017competing, hagiu2020data}, but typically model learning as a reduced-form function of past consumer history and focus on the role of prices, as opposed to the reputational consequences of learning.

\cite{schmalensee1982product, bagwell1990informational} investigate how buyer uncertainty about product quality can serve as a barrier to entry for late arrivers. We observe a similar effect when we investigate the role that reputation can serve as a barrier to entry. However, in our model this effect is further strengthened by the fact that the firms have to learn while competing adding that the incumbent may not only have a reputational advantage, but additionally has a further advantage due to the data it acquires in the incumbency period. Thus, our model also highlights the role that data can serve as a barrier to entry in online markets which has similarly been noted in \cite{lambrecht2015can, de2020data}.}

Finally, relaxed versions of rationality similar to ours are found in several notable lines of work. For example, ``random agents" (a.k.a. noise traders) can side-step the ``no-trade theorem'' \citep{Milgrom-Stokey-82}, a famous impossibility result in financial economics. The \SoftMaxRandom model is closely related to the literature on \emph{product differentiation}, starting from \cite{Hotelling-29}, see \cite{Perloff-Salop-85} for a notable later paper. There is a large literature on non-existence of equilibria due to small deviations   (which is related to the corresponding result for \HardMaxRandom), starting with \cite{Rothschild-Stiglitz-76} in the context of health insurance markets. Notable recent papers \citep{Veiga-Weyl-16,Azevedo-Gottlieb-17} emphasize the distinction between \HardMax and versions of \SoftMaxRandom.

### 2020-07-08 journal paper/content/sec-intro.tex
updated related work + intro

\xhdr{Economic interpretation: anti-trust perspective.}
Data advantage is significant as a possible barrier to entry in the digital economy
\citep[\eg][]{jullien2019economics,de2020data, hagiu2020data}.
Our model directly speaks to this literature as we endogenize data-driven network effects by explicitly considering the dynamics of learning under competition. Our results are similar to those on ``market tipping" because of indirect network effects \ascomment{some citations??}, and show channels through which differences in data ownership can result in market tipping. We observe two important differences compared to prior work:

\begin{itemize}
\item What matters is not just the quantity of data but its quality: how informative it is for the learning problem at hand. Data collected by better bandit algorithms tends to be more informative.

    \ascomment{Guy: cite prior work that only talks about data quantity? Are we really the first paper on anti-trust that talks about data quality? Seems like a big claim, we better be right on this. }

\item The ``amplification effect" of the data advantage runs contrary to prior work on learning without competition \citep{varian2018artificial,lambrecht2015can,bajari2018impact}, which found that small amounts of additional data do not provide significant improvement in eventual outcomes.
\end{itemize}

\noindent We conclude that competition dynamics are pertinent for this discussion.

>>>>

\xhdr{Economic interpretation: data network effects.}
\gaedit{Our model speaks to the ongoing debate on the role that data plays in the digital economy. One fundamental question in this debate is whether data can serve a similar role as traditional network effects, whereby competition for the market is more important than competition in the market due to the winner-take-all nature of such markets \citep{Rysman09, jullien2019economics}. The ``death spiral" effect that we find has a similar flavor as this as once a firm engages in exploration it gets completely starved of users. Indeed, we further find that a small data advantage for one firm gets amplified under competition and leads to that firm taking the entire market. A distinctive feature of our approach is that we explicitly model the learning problem of the firms and consider them as deploying algorithms for solving this problem.  Thus, we do not explicitly model such network effects, but rather they arise endogenously from our setup.

Our results highlight that understanding the performance of learning algorithms in isolation does not necessarily translate to understanding their impact in competition, precisely due to the fact that competition leads to the endogenous generation of data observed by the firms. Approaches such as \cite{bajari2018impact} that argue that the diminishing returns to scale and scope of data in isolation render such data feedback loops as non-existent ignore the differences induced by learning in isolation versus under competition. Furthermore, explicitly incorporating the interaction between learning technology and data creation allows us to speak on how data advantages are characterized and amplified not only by data quantity, but also the increased data quality gathered by better learning algorithms.
}
